---
title: "Canada_Border"
author: "Min-Yi Chen"
date: "2023-11-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# load("~/Documents/ISU/23Fall/STAT579/final project/data/border_clean.Rdata")
load("C:/Users/rebek/github/packages/borderxing/data/border_clean.Rdata")
library(tidyverse)
library(plotly)
library(gridExtra)
#ggplotly()
#read data
Canada_border <- border %>%
  filter(Border == "US-Canada Border")%>%
  mutate(
    Year = factor(Year,levels = c(1996,1997,1998,1999,2000,2001,2002,
                                  2003,2004,2005,2006,2007,2008,2009,2010,2011,
                                  2012,2013,2014,2015,2016,2017,2018,2019,2020,
                                  2021,2022,2023))
  )
```

#port code
How many port have complete set
How many port should we consider ?
Does Port_name change ? There's 118 levels but I only have 90 rows.
Only 89 when I group_by Port_name

#Eastport
```{r}
Canada_border %>%
  filter(Port_name == 'Eastport')%>%
  group_by(Port_code)%>%
  summarise(
    n = n()
  )
```
# Eastport has two Port_code
```{r}
Canada_border %>%
  filter(Port_name == 'Eastport')%>%
  group_by(Port_code)%>%
  ggplot(aes(x = Year,y = Value,colour = Port_code))+geom_point()
```

```{r}
Canada_border %>%
  filter(Port_name == 'Eastport')%>%
  group_by(Port_code)%>%
  ggplot(aes(x = Year,y = Value,colour = Port_code))+geom_boxplot()
```
# State
```{r}
Canada_border %>%
  group_by(State) %>%
  ggplot(aes(x = Date, y=Value))+geom_line()+facet_wrap(~State)

Canada_border %>%
  group_by(State) %>%
  ggplot(aes(x = Date, y=Value))+geom_line()+facet_wrap(~State,scales = 'free_y')

```

#Measure
```{r}
Canada_border %>%
  group_by(Measure) %>%
  ggplot(aes(x = Date, y=Value))+geom_line()+facet_wrap(~Measure)

Canada_border %>%
  group_by(Measure) %>%
  ggplot(aes(x = Date, y=Value))+geom_line()+facet_wrap(~Measure,scales = 'free_y')

```

We can see some seasonal patterns in measure

# How about group_by State with specific measure ?

#Bus
```{r}
Canada_border %>%
  filter(Measure == 'Buses')%>%
  group_by(State) %>%
  ggplot(aes(x = Date, y=Value))+geom_line()+facet_wrap(~State)
#montana Bus
Canada_border %>%
  filter(Measure == 'Buses')%>%
  group_by(State) %>%
  ggplot(aes(x = Date, y=Value))+geom_line()+facet_wrap(~State,scales = 'free_y')

```

#Personal Vehicles
```{r}
Canada_border %>%
  filter(Measure == 'Personal Vehicles')%>%
  group_by(State) %>%
  ggplot(aes(x = Date, y=Value))+geom_line()+facet_wrap(~State)

Canada_border %>%
  filter(Measure == 'Personal Vehicles')%>%
  group_by(State) %>%
  ggplot(aes(x = Date, y=Value))+geom_line()+facet_wrap(~State,scales = 'free_y')

```

#Personal Vehicle Passengers
```{r}
Canada_border %>%
  filter(Measure == 'Personal Vehicle Passengers')%>%
  group_by(State) %>%
  ggplot(aes(x = Date, y=Value))+geom_line()+facet_wrap(~State)

Canada_border %>%
  filter(Measure == 'Personal Vehicle Passengers')%>%
  group_by(State) %>%
  ggplot(aes(x = Date, y=Value))+geom_line()+facet_wrap(~State,scales = 'free_y')
```

I think Personal Vehicles is a factor.

```{r}
Among_State <- Canada_border %>%
  group_by(State) %>%
  ggplot(aes(x = Date, y=Value))+geom_line()+facet_wrap(~State,scales = 'free_y')
Bus_Among_State <- Canada_border %>%
  filter(Measure == 'Buses')%>%
  group_by(State) %>%
  ggplot(aes(x = Date, y=Value))+geom_line()+facet_wrap(~State,scales = 'free_y')

Personal_Vehicles_Among_State <- Canada_border %>%
  filter(Measure == 'Personal Vehicles')%>%
  group_by(State) %>%
  ggplot(aes(x = Date, y=Value))+geom_line()+facet_wrap(~State,scales = 'free_y')

Personal_Vehicles_Passengers_Among_State <- Canada_border %>%
  filter(Measure == 'Personal Vehicle Passengers')%>%
  group_by(State) %>%
  ggplot(aes(x = Date, y=Value))+geom_line()+facet_wrap(~State,scales = 'free_y')

grid.arrange(Among_State, Bus_Among_State, nrow = 2)
```
```{r}
grid.arrange(Among_State, Personal_Vehicles_Among_State, nrow = 2)
```

```{r}
grid.arrange(Among_State, Personal_Vehicles_Passengers_Among_State, nrow = 2)
```

#Consider average in each month
```{r}
Canada_border %>%
  group_by(Year,Month) %>%
  summarise(
    average_value = mean(Value,na.rm = T)
  ) %>%
  ggplot(aes(x = factor(Month), y=average_value,group = Year,colour = Year))+geom_line()

```

```{r}
Canada_border %>%
  group_by(Year,Month) %>%
  summarise(
    average_value = mean(Value,na.rm = T)
  ) %>%
  ggplot(aes(x = factor(Month), y=average_value,colour = Year))+geom_line(aes(group = Year)) +facet_wrap(~Year)+
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
```



