---
title: "US Border X-ing"
subtitle: "<br/> Stat 579 Final Project | Iowa State University"
author: "Min-Yi Chen, Liam Hemingway, Rebekah Scott, Zihao Yang"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---



```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```


```{r, include = FALSE}
# load the data with the here package, more at https://here.r-lib.org/
here::i_am("presentation_minyi.Rmd") # establish where this document is
# here::here("../../data/border_clean.Rdata") # make a path to where the data is
#load(here::here("../data/border_clean.Rdata")) # load the data
load("~/Documents/ISU/23Fall/STAT579/final project/data/border_clean.Rdata")
library(tidyverse)
library(plotly)
library(gridExtra)
#ggplotly()
#read data
Canada_border <- border %>%
  filter(Border == "US-Canada Border")%>%
  mutate(
    Year = factor(Year,levels = c(1996,1997,1998,1999,2000,2001,2002,
                                  2003,2004,2005,2006,2007,2008,2009,2010,2011,
                                  2012,2013,2014,2015,2016,2017,2018,2019,2020,
                                  2021,2022,2023)),
    date=lubridate::my(Date),
         month=month(date),
         year=year(date)
  )
```


<!-- Image credit: Rebekah Scott -->


---
class: inverse, center, middle

# US-Canada Border - Min-Yi

---

# Bus Transporation in US-Canada Border

Outliers : 

- Nov 1999, Port Sweetgrass in Montana

- June 2021, Port Limestone in Maine

```{r,echo=FALSE,out.width="50%",fig.align='center'}
Canada_border %>%
  filter(Measure == 'Buses' & Type == 'Object')%>%
  group_by(State) %>%
  ggplot(aes(x = year + (month-1)/12, y=Value))+geom_point(colour = "deepskyblue4")+facet_wrap(~State,scales = 'free_y') +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
  geom_point(data=Canada_border %>%
             filter(Measure == 'Buses' & Type == 'Object' & State == 'Montana' & Value == 1375),
             pch = 24,
             size=4, 
             colour = "red") + geom_point(data=Canada_border %>%
             filter(Measure == 'Buses' & Type == 'Object' & State == 'Maine' & Value == 173),
             pch = 24,
             size=4, 
             colour = "red")+scale_color_brewer(type = "seq", palette = "Blues") + theme_bw() + 
  xlab("Date") +
  ylab("Number of Crossing") + ggtitle("Numbers of Crossing of Object by State")
```

---

Analysis of the total object transportation for the year 1999 in Montana and the year 2021 in Maine. 

These pinpoint data points (outliers) do not align with the peak of the plot.

```{r all-trans-object, fig.cap = "Plot of Transporation via object by Month",echo=FALSE,fig.align='center',out.width="50%"}

Montana_1999<- Canada_border %>%
  filter(Year == '1999' & State == 'Montana' & Type == 'Object')%>%
  group_by(Month) %>%
  summarise(
    Transporation_via_object = sum(Value)
  )

Maine_2021<- Canada_border %>%
  filter(Year == '2021' & State == 'Maine' & Type == 'Object')%>%
  group_by(Month) %>%
  summarise(
    Transporation_via_object = sum(Value)
  )

ggplot() + 
  geom_line(data = Montana_1999,aes(x = Month, y=Transporation_via_object,group = 1,colour = 'Montana_1999')) + 
  geom_line(data = Maine_2021,aes(x = Month, y=Transporation_via_object,group = 1,colour = 'Maine_2021'))+
  scale_color_manual(name = "Year", values = c("Montana_1999" = "deepskyblue3", "Maine_2021" = "deepskyblue4"))+geom_point(aes(x='Jun',y=70976),colour="deepskyblue4")+geom_point(aes(x='Nov',y=70074),colour="deepskyblue3") + theme_bw() + ylab('Numbers of Crossing with Obejct') 

```

---

# Bus Passengers Transporation in US-Canada Border

Outliers : 

- 46712 Bus Passengers in November 1998 in Maine at Port Jackman

- 159086 Bus Passengers in October 159086 in Washington at Port Blaine

- 6026 Bus Passengers in June 1997 in Minnesota at Port Noyes

- 2203 Bus Passengers in May 2000 in Idaho at Eastport ID

- 135753 Bus Passengers in December 1997 in Michigan at Detroit

- 19322 Bus Passengers in May 2019 in Vermont at Highgate Springs

---

```{r bus-ppl, echo=FALSE,out.width="80%",fig.align='center'}
Canada_border %>%
  filter(Measure == 'Bus Passengers' & Type == 'People')%>%
  group_by(State) %>%
  ggplot(aes(x = year + (month-1)/12, y=Value))+geom_point(colour = "deepskyblue4")+facet_wrap(~State,scales = 'free_y')+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
  geom_point(data=Canada_border %>%
             filter(Measure == 'Bus Passengers' & Type == 'People' & State == 'Maine' & Value == 46712 & Year == '1998'),
             pch = 24,
             size=4, 
             colour = "red")+
  geom_point(data=Canada_border %>%
             filter(Measure == 'Bus Passengers' & Type == 'People' & State == 'Washington' & Value == 159086 & Year == '2006'),
             pch = 24,
             size=4, 
             colour = "red")+
  geom_point(data=Canada_border %>%
             filter(Measure == 'Bus Passengers' & Type == 'People' & State == 'Minnesota' & Value == 6026 & Year == '1997'),
             pch = 24,
             size=4, 
             colour = "red")+
  geom_point(data=Canada_border %>%
             filter(Measure == 'Bus Passengers' & Type == 'People' & State == 'Idaho' & Value == 2203 & Year == '2000'),
             pch = 24,
             size=4, 
             colour = "red")+
  geom_point(data=Canada_border %>%
             filter(Measure == 'Bus Passengers' & Type == 'People' & State == 'Michigan' & Value == 135753 & Year == '1997'),
             pch = 24,
             size=4, 
             colour = "red")+
  geom_point(data=Canada_border %>%
             filter(Measure == 'Bus Passengers' & Type == 'People' & State == 'Vermont' & Value == 19322 & Year == '2019'),
             pch = 24,
             size=4, 
             colour = "red") + theme_bw()+ 
  xlab("Date") +
  ylab("Number of Crossing")

```

---

We observed that the identified outliers in both Buses and Bus Passengers transportation did not coincide with the peak of the plot.


```{r ,echo=FALSE}

Maine_1998 <- Canada_border %>%
  filter(Year=='1998' & State == 'Maine' & Type == 'People') %>%
  group_by(Month) %>%
  summarise(
    Transporation_via_people = sum(Value)
  )

Washington_2006 <- Canada_border %>%
  filter(Year=='2006' & State == 'Washington' & Type == 'People') %>%
  group_by(Month) %>%
  summarise(
    Transporation_via_people = sum(Value)
  )

Minnesota_1997 <- Canada_border %>%
  filter(Year=='1997' & State == 'Minnesota' & Type == 'People') %>%
  group_by(Month) %>%
  summarise(
    Transporation_via_people = sum(Value)
  )

Idaho_2000 <- Canada_border %>%
  filter(Year=='2000' & State == 'Idaho' & Type == 'People') %>%
  group_by(Month) %>%
  summarise(
    Transporation_via_people = sum(Value)
  )

Michigan_1997 <- Canada_border %>%
  filter(Year=='1997' & State == 'Michigan' & Type == 'People') %>%
  group_by(Month) %>%
  summarise(
    Transporation_via_people = sum(Value)
  )

Vermont_2019 <- Canada_border %>%
  filter(Year=='1998' & State == 'Vermont' & Type == 'People') %>%
  group_by(Month) %>%
  summarise(
    Transporation_via_people = sum(Value)
  )

```

```{r all-trans-ppl, echo=FALSE,out.width="70%",fig.align='center'}
ggplot() + 
  geom_line(data = Maine_1998,aes(x = Month, y=Transporation_via_people,group = 1,colour = 'Maine_1998')) + 
  geom_line(data = Washington_2006,aes(x = Month, y=Transporation_via_people,group = 1,colour = 'Washington_2006')) + 
  geom_line(data = Minnesota_1997,aes(x = Month, y=Transporation_via_people,group = 1,colour = 'Minnesota_1997')) +
  geom_line(data = Idaho_2000,aes(x = Month, y=Transporation_via_people,group = 1,colour = 'Idaho_2000' )) +
  geom_line(data = Michigan_1997,aes(x = Month, y=Transporation_via_people,group = 1,colour = 'Michigan_1997' )) +
  geom_line(data = Vermont_2019,aes(x = Month, y=Transporation_via_people,group = 1,colour = 'Vermont_2019' )) +
  scale_color_manual(name = "Year", values = c("Maine_1998" = "deepskyblue1", "Washington_2006" = "deepskyblue2","Minnesota_1997" = "deepskyblue3", "Idaho_2000" = "deepskyblue4","Michigan_1997" = "steelblue","Vermont_2019" = "darkblue")) +geom_point(aes(x='Nov',y=681595),colour="deepskyblue1")+geom_point(aes(x='Oct',y=1092705),colour="deepskyblue2")+geom_point(aes(x='Jun',y=331095),colour="deepskyblue3") +geom_point(aes(x='May',y=46723),colour="deepskyblue4")+geom_point(aes(x='Dec',y=2193053),colour="steelblue")+geom_point(aes(x='May',y=281528),colour="darkblue") + theme_bw()+ 
  xlab("Date") +
  ylab("Number of Crossing")
```

---

# Seasonal Treand

By aggregating data into seven-year intervals, examining the average transportation volumes for both objects and people. We found :

- Seasonal Trend

- Reach it's zenith - Translational decline - Bounced back

```{r,echo=FALSE}
Group_1_Ob <- Canada_border %>%
  filter((Year=='1996'|Year=='1997'|Year=='1998'|Year=='1999'|Year=='2000'|Year=='2001'|Year=='2002')&Type == 'Object' ) %>%
  group_by(Month) %>%
  summarise(
    average_value = mean(Value,na.rm = T)
  )

Group_2_Ob <- Canada_border %>%
  filter((Year=='2003'|Year=='2004'|Year=='2005'|Year=='2006'|Year=='2007'|Year=='2008'|Year=='2009')&Type == 'Object' ) %>%
  group_by(Month) %>%
  summarise(
    average_value = mean(Value,na.rm = T)
  )

Group_3_Ob <- Canada_border %>%
  filter((Year=='2010'|Year=='2011'|Year=='2012'|Year=='2013'|Year=='2014'|Year=='2015'|Year=='2016')&Type == 'Object' ) %>%
  group_by(Month) %>%
  summarise(
    average_value = mean(Value,na.rm = T)
  )

Group_4_Ob <- Canada_border %>%
  filter((Year=='2017'|Year=='2018'|Year=='2019'|Year=='2020'|Year=='2021'|Year=='2022'|Year=='2023')&Type == 'Object' ) %>%
  group_by(Month) %>%
  summarise(
    average_value = mean(Value,na.rm = T)
  )

```
```{r,echo=FALSE}

avg_obj <- ggplot() + 
  geom_line(data = Group_1_Ob,aes(x = Month, y=average_value,group = 1,colour = '1996-2002')) + 
  geom_line(data = Group_2_Ob,aes(x = Month, y=average_value,group = 1,colour = '2003-2009')) + 
  geom_line(data = Group_3_Ob,aes(x = Month, y=average_value,group = 1,colour = '2010-2016')) +
  geom_line(data = Group_4_Ob,aes(x = Month, y=average_value,group = 1,colour = '2017-2023' )) +
  scale_color_manual(name = "Year Group", values = c("1996-2002" = "deepskyblue1", "2003-2009" = "deepskyblue2","2010-2016" = "deepskyblue3", "2017-2023" = "deepskyblue4")) + ggtitle('Average of crossing in object within 7 years') + ylab('Numbers of Crossing in Object')

```



```{r,echo=FALSE}
Group_1_People <- Canada_border %>%
  filter((Year=='1996'|Year=='1997'|Year=='1998'|Year=='1999'|Year=='2000'|Year=='2001'|Year=='2002')&Type == 'People' ) %>%
  group_by(Month) %>%
  summarise(
    average_value = mean(Value,na.rm = T)
  )

Group_2_People <- Canada_border %>%
  filter((Year=='2003'|Year=='2004'|Year=='2005'|Year=='2006'|Year=='2007'|Year=='2008'|Year=='2009')&Type == 'People' ) %>%
  group_by(Month) %>%
  summarise(
    average_value = mean(Value,na.rm = T)
  )

Group_3_People <- Canada_border %>%
  filter((Year=='2010'|Year=='2011'|Year=='2012'|Year=='2013'|Year=='2014'|Year=='2015'|Year=='2016')&Type == 'People' ) %>%
  group_by(Month) %>%
  summarise(
    average_value = mean(Value,na.rm = T)
  )

Group_4_People <- Canada_border %>%
  filter((Year=='2017'|Year=='2018'|Year=='2019'|Year=='2020'|Year=='2021'|Year=='2022'|Year=='2023')&Type == 'People' ) %>%
  group_by(Month) %>%
  summarise(
    average_value = mean(Value,na.rm = T)
  )

```
```{r avg-trans, echo=FALSE,out.width="50%",fig.align='center'}

avg_ppl <- ggplot() + 
  geom_line(data = Group_1_People,aes(x = Month, y=average_value,group = 1,colour = '1996-2002')) + 
  geom_line(data = Group_2_People,aes(x = Month, y=average_value,group = 1,colour = '2003-2009')) + 
  geom_line(data = Group_3_People,aes(x = Month, y=average_value,group = 1,colour = '2010-2016')) +
  geom_line(data = Group_4_People,aes(x = Month, y=average_value,group = 1,colour = '2017-2023' )) +
  scale_color_manual(name = "Year Group", values = c("1996-2002" = "deepskyblue1", "2003-2009" = "deepskyblue2","2010-2016" = "deepskyblue3", "2017-2023" = "deepskyblue4")) + ggtitle('Average of crossing in people within 7 years')+ ylab('Numbers of Crossing in People')

gridExtra::grid.arrange( avg_obj, avg_ppl, nrow = 2, ncol = 1)

```


