---
title: "rmarkdown"
author: "Zihao Yang"
date: "`r Sys.Date()`"
output: bookdown::html_document2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r warning=FALSE}
setwd("/Users/yangzihao/Desktop/Project/STAT 579/finalproject2/data")
load("border_clean.Rdata")
```

```{r include=FALSE}
library(tidyverse)
library(plotly)
library(gridExtra)
library(ggplot2)
library(bookdown)
```
```{r}
Canada_border <- border %>%
  filter(Border == "US-Canada Border")%>%
  mutate(
    Year = factor(Year,levels = c(1996,1997,1998,1999,2000,2001,2002,
                                  2003,2004,2005,2006,2007,2008,2009,2010,2011,
                                  2012,2013,2014,2015,2016,2017,2018,2019,2020,
                                  2021,2022,2023))
  )
Mexico_border <- border %>%
  filter(Border == "US-Mexico Border")%>%
  mutate(
    Year = factor(Year,levels = c(1996,1997,1998,1999,2000,2001,2002,
                                  2003,2004,2005,2006,2007,2008,2009,2010,2011,
                                  2012,2013,2014,2015,2016,2017,2018,2019,2020,
                                  2021,2022,2023))
  )
```

# Time series

```{r tsMexico, echo=F, message=F, fig.cap="Time series data related to the value of Mexico border."}
Mexico_value <- Mexico_border %>%
  filter(Type == "People")%>%
  group_by(Month,Year)%>%
  summarize(Value = sum(Value))
Mexico_value$Date <- paste(Mexico_value$Year, Mexico_value$Month, "01", sep = "-")
Mexico_value$Date <- as.Date(Mexico_value$Date,format = "%Y-%b-%d")
Mexico_value$Date <- Mexico_value[order(Mexico_value$Date, decreasing = FALSE), ]
Mexico_value %>%
  ggplot(aes(x = Date$Date, y = Value)) +
  geom_line(color = "skyblue2") +
  theme_bw()+
  labs(title = "Value vs. Date",
       x = "Date",
       y = "Value") 
Mexico_time_series <- ts(data = Mexico_value$Value, frequency = 12)
Mexico_decomposed_series <- decompose(Mexico_time_series, type = "additive")
Mexico_season <- Mexico_decomposed_series$seasonal[1:12]
year <- 0
start_date <- as.Date(paste(year, "-01-01", sep = ""))
end_date <- as.Date(paste(year, "-12-31", sep = ""))
dates <- seq(start_date, end_date, by = "month")
a <- data.frame(month1 = dates, Mexico_value = Mexico_season)
a$month1 <- as.Date(a$month1)
```

Figure \@ref(fig:tsMexico) shows the value of the Mexico border vs time. We utilize the 'additive' method to decompose the population data at the Mexico border, dividing it into three components: trend, season, and random parts. Our focus for this project lies in analyzing the seasonal component.

```{r tsCanada, echo=F, message=F, fig.cap="Time series data related to the value of Canada border."}
Canada_value <- Canada_border %>%
  filter(Type == "People")%>%
  group_by(Month,Year)%>%
  summarize(Value = sum(Value))

Canada_value$Date <- paste(Canada_value$Year, Canada_value$Month, "01", sep = "-")
Canada_value$Date <- as.Date(Canada_value$Date,format = "%Y-%b-%d")
Canada_value$Date <- Canada_value[order(Canada_value$Date, decreasing = FALSE), ]
Canada_value %>%
  ggplot(aes(x = Date$Date, y = Value)) +
  geom_line(color = "skyblue2") +
  theme_bw()+
  labs(title = "Value vs. Date",
       x = "Date",
       y = "Value") 
Canada_time_series <- ts(data = Canada_value$Value, frequency = 12)
Canada_decomposed_series <- decompose(Canada_time_series, type = "additive")
Canada_season <- Canada_decomposed_series$seasonal[1:12]
```
Figure \@ref(fig:tsCanada) shows the value of the Canada border vs time. We utilize the 'additive' method to decompose the population data at the Canada border, dividing it into three components: trend, season, and random parts. Our focus for this project lies in analyzing the seasonal component.

```{r tscompare, echo=F, message=F, fig.cap="Comparison between the seasonal components of two borders."}
year <- 0
start_date <- as.Date(paste(year, "-01-01", sep = ""))
end_date <- as.Date(paste(year, "-12-31", sep = ""))
dates <- seq(start_date, end_date, by = "month")
b<- data.frame(Canada_value = Canada_season)
c<- cbind(a,b)
library(reshape2)
c_long <- melt(c, id.vars = "month1")
c_long%>%
  ggplot(aes(x = month1, y = value, color = variable)) +
  geom_line() +
  scale_fill_brewer(type = "seq",palette = "Blues")+
  theme_bw()+
  labs(x = "month", y = "Value", color = "Data Series") +
  ggtitle("Two Sets of Values on Same Plot")
```

Observing the Figure \@ref(fig:tscompare), it's evident that the seasonal patterns of Mexico and Canada differ significantly. The observed differences in seasonal part between Mexico and Canada border crossings may caused by various factors, including cultural, economic, and climatic influences. The distinct peaks in border crossing values during different months suggest varying trends in travel behavior and trade activities between the two countries. Mexico's peak occurs in November, whereas Canada's peak is in May. This discrepancy could potentially be attributed to weather effects; May might be a comfortable month for people to travel or conduct business through the Canada border, whereas November remains warm along the Mexico border.

# Big event influence

```{r impact911, echo=F, message=F, fig.cap= "Impact of the 9/11 event on the Mexico and Canada borders. The darkblue line represents the value of September in the year before the event, while the olivegreen line represents the value of September in the year after the event."}
# 911
start_date1 <- ymd("2000-01-01")  # Convert the start date to a Date object
end_date1 <- ymd("2003-01-01")    # Convert the end date to a Date object
date_range1 <- seq(start_date1, end_date1, by = "day")  

p1 <- border %>% 
  mutate(date=lubridate::my(Date),
         month=month(date),
         year=year(date)) %>% 
  filter(date %in% date_range1 )%>%
  group_by(month,year,Border) %>% 
  summarize(Value=sum(Value)) %>% 
  ggplot(aes(x=year + (month-1)/12,y=Value))+
  geom_line(color = "skyblue2")+
  geom_vline(xintercept=2001+(9-1)/12,color='bisque3')+ # sep 11
  facet_wrap(~Border,scales = "free_y")+
  theme_bw()+
  xlab("Time")+
  theme(legend.position = "none")

line1 <- data.frame(Border = c("US-Canada Border","US-Mexico Border"),wt = c(13369583,32002624))
line2 <- data.frame(Border = c("US-Canada Border","US-Mexico Border"),wt = c(10848047,28613279))
p2 <- p1 + geom_hline(aes(yintercept = wt),line1, color = 'skyblue4')
p2 + geom_hline(aes(yintercept = wt),line2, color = 'darkolivegreen3')+
  ggtitle("911")
```

As observed in the Figure \@ref(fig:impact911), the 9/11 event had a noticeable impact on the value of both borders. There was a significant decline in value on September 11th, 2001, followed by a subsequent recovery the following year.

```{r impactcovid, echo=F, message=F, fig.cap="The impact of the COVID-19 event on the Mexico and Canada borders. The darkblue line represents the value of March in the year prior to the event, while the olivegreen line represents the value of March in the subsequent year."}
#covid
start_date2 <- ymd("2019-01-01")  # Convert the start date to a Date object
end_date2 <- ymd("2022-01-01")    # Convert the end date to a Date object
date_range2 <- seq(start_date2, end_date2, by = "day")  

p3 <- border %>% 
  mutate(date=lubridate::my(Date),
         month=month(date),
         year=year(date)) %>% 
  filter(date %in% date_range2 )%>%
  group_by(month,year,Border) %>% 
  summarize(Value=sum(Value)) %>% 
  ggplot(aes(x=year + (month-1)/12,y=Value))+
  geom_line(color = "skyblue2")+
  geom_vline(xintercept=2020+(3-1)/12,color='bisque3')+ # covid
  facet_wrap(~Border,scales = "free_y")+
  theme_bw()+
  xlab("Time")+
  theme(legend.position = "none")

line3 <- data.frame(Border = c("US-Canada Border","US-Mexico Border"),wt = c(7098470,23889091))
line4 <- data.frame(Border = c("US-Canada Border","US-Mexico Border"),wt = c(1818274,16060812))
p4 <- p3 + geom_hline(aes(yintercept = wt),line3, color = 'skyblue4')
p4 + geom_hline(aes(yintercept = wt),line4, color = 'darkolivegreen3')+
  ggtitle("Covid")

```

From the Figure \@ref(fig:impactcovid), it's evident that COVID-19 had a considerable impact on the values of both borders, notably declining sharply in March 2020. This event not only affected the immediate values but also influenced the seasonal patterns of each border. The values remained consistently low in the subsequent periods following the initial decline.

# Comparison 

```{r include=FALSE}
Mexico_border_max10 <- Mexico_border %>%
  filter(Type == "People") %>%
  slice_max(Value,n = 10)

Canada_border_max10 <- Canada_border %>%
  filter(Type == "People") %>%
  slice_max(Value,n = 10)
factor(Mexico_border_max10$Port_name)
factor(Canada_border_max10$Port_name)
```

We selected the Mexico and Canada border data, filtering by the 'Type' attribute equal to 'People.' From this dataset, we identified the top 10 largest values and extracted their corresponding 'Port_name.' The findings revealed that for the Mexico border, the predominant port is El Paso, while for the Canada border, the prominent ports are Buffalo Niagara Falls and Detroit. As these three ports exhibit significant interest, we will conduct further research on them.

```{r}
border1 <- border
border1$Date <- as.Date(paste0("01 ", border1$Date), format = "%d %b %Y")
```
```{r bar, echo=F, message=F, fig.cap="Comparison object"}
border1 %>%
  filter(Port_name =="El Paso" |Port_name =="Detroit"|Port_name =="Buffalo Niagara Falls")%>%
  filter(Type == "Object")%>%
  group_by(Port_name, Measure) %>%
  summarize(Value=sum(Value))%>%
  ggplot(aes(x = factor(Port_name),y = Value))+
  geom_bar(stat = "identity",position = position_dodge(),color = "skyblue3",fill = "skyblue3") +
  scale_fill_brewer(palette = "Blues")+
  theme_bw()+
  facet_wrap(~Measure, scales = "free_y")+
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

In Figure \@ref(fig:bar) we compare the three largest ports identified earlier, there are notable differences emerge between Mexico and Canada. It appears that individuals crossing the Mexico border tend to prefer personal vehicles, whereas those crossing the Canada border favor public transportation. Additionally, we observed that the value of empty rail and truck crossings is higher at the Mexico border than at the Canada border, indicating a lower utilization rate of traffic infrastructure in Mexico. Addressing this discrepancy is crucial when considering environmental impact and energy efficiency.